---
title: "Introduction to BinaryClust2 data analysis pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BinaryClust2 data analysis pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
#Introduction to BinaryClust2 data analysis pipeline
BinaryClust2 was further developed based on BinaryClust with adaption of the framework and structure of CyTOF data analysis workflow described by Nowicka et al. This package enables automatic classification and annotation of main cell types as well as in-depth interrogation of certain sub-population of interest.

After data quality check, batch effect evaluation and correction, here we can continue for differential analysis. 

First, load below R packages. 
```{r setup}
#load required R packages 
library(BinaryClust2)
library(BinaryClust2)
library(CATALYST)
library(readxl)
library(flowCore)
library(SingleCellExperiment)
library(cowplot)
library(ggplot2)
library(reshape2)
library(dplyr)
library(caret)
library(rstatix)
library(Rphenograph)
library(readr)
library(RColorBrewer)
library(igraph)
```
#Data Preparation
Users are required to prepare two excel files of panel metadata and sample metadata. Briefly, panel metadata describes the panel used in the experiment, which includes three columns: ‘fcs_colname’, ‘antigen’ and ‘marker_class’. ‘fcs_colname’ is metal/channel names and can be checked via either cytobank or using flowCore package function: colnames(); ‘Antigen’ refers to marker names which can be checked from the applied CyTOF kit instructions; Based on CATALYST, markers can be classified as ‘type’, which means lineage markers and will be used for clustering and ‘state’ which indicates functional markers like intracellular signalling proteins etc. Sample metadata indicates the experimental design which contains the information of fcs file name(‘file_name’), ‘sample_id’, ‘patient_id’, experimental grouping(‘condition’), or additional information like ‘outcome’ etc. Please prepare you own sample metadata accordingly. 

Before importing into R-based analytical pipelines, all fcs files are supposed to be normalized for signal fluctuation during acquisition and cleaned up to exclude all sorts of debris, doublets etc. using third-party platform, for instance, Cytobank(https://premium.cytobank.org/cytobank/login). Some users might prefer to do compensation dependent on their preference and previous experience of handling certain types of panel and samples. 

Example dataset used here can be downloaded from Zenodo(https://doi.org/10.5281/zenodo.7982165).
```{r setup}
#set file path for fcs files, sample metadata and panel metadata
fcs_dir <- getwd()
metadata_dir <- 'metadata2.xlsx'
panel_dir <- "panel_metadata2.xlsx"
md <- read_excel(metadata_dir)
panel <- read_excel(panel_dir)
#read fcs files as a flowset
fcs_files <- list.files(fcs_dir, pattern = "fcs$")
fs <- read.flowSet(file.path(fcs_dir, fcs_files))
#Data preparation
#check if all the colname in panel file is in the flowset colames
all(panel$fcs_colname %in% colnames(fs))
#set levels for the columns of metadata file
md$condition <- factor(md$condition, levels = c("JAK2", "Non_JAK2"))
md$sample_id <- factor(md$sample_id,
                       levels = md$sample_id[order(md$condition)])
md$patient_id <- factor(md$patient_id)
md$outcome <- factor(md$outcome, levels= c('1','2','3'))
#construct a sce object
sce <- prepData(fs, panel, md, md_cols = list(file='file_name',id='sample_id',factors=c('outcome','condition','patient_id')), features = panel$fcs_colname)
#check the number of total cells;if the data is too heavy then downsampling could be considered 
dim(sce)
#example for how to perform downsampling, then define the downsampled data as sce_d 
sce_d <-downsample(sce,n=2000,seed=1234)
#check cell number of the downsampled sce object 
dim(sce_d)
```
Or alternatively, you can download the defined R objects directly into memory to skip the above import process, as shown below: 
```{r}
#download sample metadata, panel metadata and the fcs object from Zenodo
options(timeout = 200)
load(url("https://zenodo.org/record/7966205/files/md.RData?download=1"))
head(md)
load(url("https://zenodo.org/record/7966205/files/panel.RData?download=1"))
head(panel)
load(url("https://zenodo.org/record/7966205/files/MPN_sce.RData?download=1"))
```
#Binary classification and annotation 
The next is to use the key function of BinaryClust2 to automatically classify markers and cell populations based on user pre-defined classification matrix. Hence, please prepare a csv file(here is named as 'types' in the example), where user can define each cell population based on different maker expression and their gating strategy. ‘+’ means positively expressed, ‘-’ means negatively expressed whereas ‘A’ means ‘any’. The specific names of the marker listed in the types file shall be exact the same as panel metadata. 
```{r}
#import user-defined types file
class_dir <- 'types.csv'
types <- read_csv('types.csv')
#check the behavior of specific marker distribution  
displayMarkers(sce,types,color_by='sample_id',ncol=3)
#perform binary classification
exprs <-t(assay(sce, 'exprs'))
binary.results <- binaryClass(exprs, class_dir)
sce$cluster_id <- unlist(binary.results)
#check binary classification results of all cells
plotbcFreq(sce,binary.results)
plotbcHeatmap(sce,binary.results,remove.unclass=F)
#plot DR colored by binary classification results
sce <- runDR(sce, dr = "UMAP", cells = 2000, features = "type")
sce <- runDR(sce, dr = "TSNE", cells = 2000, features = "type")
plotDR(sce, "UMAP", color_by = "cluster_id")
plotDR(sce, "TSNE", color_by = "cluster_id")
```
# Binary classification results: differential analysis visualisation(cell frequency and state 
Then plot the percentage/composition of the various cell lineages in different study groups.In this example dataset, we have two factors determining study groups: 'condition'(JAK2/non_JAK2) and ‘outcome’(1/2/3).  
```{r}
#plot DR facet by 'condition'
plotDR(sce, "UMAP", color_by = "cluster_id",facet_by='condition')
plotDR(sce, "TSNE", color_by = "cluster_id",facet_by='condition')
#plot DR facet by 'outcome'
plotDR(sce, "UMAP", color_by = "cluster_id",facet_by='outcome')
plotDR(sce, "TSNE", color_by = "cluster_id",facet_by='outcome')
#plot stacked histogram and boxplot comparing different study groups
plotFreq(sce,type='stacked',group_by='condition')
plotFreq(sce,type='box',group_by='condition')
plotFreq(sce,type='stacked',group_by='outcome')
plotFreq(sce,type='box',group_by='outcome')
#plot state marker expression in each cluster
plotStateMarkerExprs(sce,group='condition')
plotStateMarkerExprs(sce,group='outcome')
```
#statistical analysis of binary classification results
Users can extract the raw data(table) for cell frequencies to modify it based on statistical rules, for instance, remove certain outliers etc., then input the table for statistics. Instead, users can go straight from sce to get the statistical results.      

For two study groups(n=2), Mann-Whitney Wilcoxon test will be used. If n>2, then Kruskal Wallis test will be applied followed by post hoc analysis. There are two options for post hoc analysis (pairwise Wilcoxon test, and Dunn’s test) with multiple testing correction using Benjamini-Hochberge Procedure(default). 
```{r}
# get raw data of cell frequencies
fq<-freqData(sce,metadata=T)
#statistical comparison for cell frequencies using sce object  
stat_outcome<-statTest(sce,group='outcome',posthoc = T, method = 'pairwise.wilcox.test')
stat_condition<-statTest(sce,group='condition')
#statistical comparison using frequency table 
stat_outcome<-statTest(abuntable=fq,group='outcome',posthoc = T, method = 'pairwise.wilcox.test')
stat_condition<-statTest(abuntable=fq,group='condition')

```
#extract cluster of interest for in-depth interrogation
After binary classification and statistical analysis, or perhaps based on literature and preliminary data, users may find one or two certain cluster are of particular interest, then decide to go deeper for more in-depth investigation. In this example, we are interested in CD4 T cells. In BinaryClust, Phenograph algorithm is available for clustering; and the rest like merging clusters, ploting hierachy clustering heatmap is compatible to CATALYST functions.    
```{r}
#subset CD4 T cell population for in-depth interrogation
CD4_sce<-extractCluster(sce,cluster='T Cells, CD4')
# run clustering algorithm Phenograph, and assign it to a new object named CD4_rph_sce
set.seed(1234)
CD4_rph_sce <-runPheno(CD4_sce,k=20,downsample=F,seed=134,type_markers = T)
plotClusterHeatmap(CD4_rph_sce,
                   hm2 = NULL, k = "rph", m = NULL,
                   cluster_anno = TRUE, draw_freqs = TRUE)

#OR users may choose flowSOM for clustering 
CD4_sce<-CATALYST::cluster(CD4_sce, features = type_markers(CD4_sce),
                        xdim = 10, ydim = 10, maxK = 40, seed = 1234)
plotClusterHeatmap(CD4_sce,
                   hm2 = NULL, k = "meta20", m = NULL,
                   cluster_anno = TRUE, draw_freqs = TRUE)

#merge clusters using CATALYST functions
merging_table1<-read_excel("cluster_merging3.xlsx")
head(data.frame(merging_table1))
CD4_rph_sce <- mergeClusters(CD4_rph_sce, k = "rph",
                     table = merging_table1, id = "merging1")
plotClusterHeatmap(CD4_rph_sce, k = "merging1")
plotAbundances(CD4_rph_sce, k = "merging1", by = "cluster_id", shape = "patient_id")
#run DR 
CD4_rph_sce <- runDR(CD4_rph_sce, dr = "UMAP", cells = 2000, features = "type")
CD4_rph_sce <- runDR(CD4_rph_sce, dr = "TSNE", cells = 2000, features = "type")
plotDR(CD4_rph_sce, "UMAP", color_by = "cluster_id",facet_by='condition')
plotDR(CD4_rph_sce, "TSNE", color_by = "cluster_id",facet_by='condition')
#extract raw frequency data
fq_CD4T<-freqData(CD4_rph_sce,metadata=T)
#statistical comparison for cell frequencies  
stat_outcome_CD4T<-statTest(CD4_rph_sce,group='outcome',posthoc = T, method = 'pairwise.wilcox.test')
stat_condition_CD4T<-statTest(CD4_rph_sce,group='condition',posthoc = T, method = 'pairwise.wilcox.test')
```

